---
title: "Computational Musicology Portfolio"
author: "Aram Elias"
date: "2/10/2021"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(spotifyr)
library(ggplot2)
```

# Analysis: Ayreon
Dutch progressive-metal band, *Ayreon*, has released six concept albums between 1995 and 2017 that are known for their unique sound. It may be worthwhile to compare these albums to one another to see how the band has developed over time.

Some songs in different albums reference one another, and for these it could also be interesting to see how much they have in common musically speaking. Some other songs are considered outliers due to not sounding the same as the rest of the album, or even having more in common with other albums to the human ear.

With most songs being over five minutes in length and having distinct musical segments with complex rhythms, and each album consisting of about fifteen songs, a large amount of data on this band is available on Spotify.

A limitation of this corpus would be a lack of other material to compare it with; however, with the band citing older bands such as Pink Floyd as inspiration, looking into similarities there might also be a good avenue for research.

```{r}
# Creating the dataset
ayreon <- get_playlist_audio_features("", "3O1z9gRNz9DiB9wujbgLiV")
ayreon <- subset(ayreon, track.album.name != "Timeline")


ayreon %>% ggplot(aes(x = energy)) + geom_histogram(binwidth = 0.1) +
 facet_wrap(~track.album.name)
```

# Overview
This plot is an overview of Ayreon's six albums.